Prefix-free graphs and suffix array construction in sublinear space

### 2023-06-23 manuscript/pfg.tex
finished pfg

\section{Prefix-free graphs}
The idea of prefix-free graphs is inspired by a technique used in the tool rsync named Context-Triggered Piecewise Hashing (CTPH) \cite{kornblum2006identifying}.
CTPH uses a rolling hash to partition a string into substrings such that long repeated substrings are partitioned the same way.
These substrings are then hashed with a traditional hash function and stored as a string signature.
The signatures of several files are then compared to determine their changes.

In prefix-free graphs, we partition each sequence in a given pangenome into \emph{segment}s.
These segments form nodes of the prefix-free graph, and their adjacencies in the original sequences constitute edges.
The sequences are then represented as \emph{path}s in the graph.

The segments have two essential characteristics, which makes them a good choice for nodes of a pangenomic graph.
Similarly to CTPH, long repeated sequences will be partitioned the same way.
Furthermore, no segment is a prefix of another, making a set of segments prefix-free.
The second characteristic is crucial for connecting prefix-free graphs and suffix arrays, as will be presented in the next section.

@@

To illustrate the entire procedure, consider a set of sequences $P = {}$ and a set of trigger words $T = {AC, CG}$.
After the partitioning, we obtain a set of segments with IDs ${}$ and a list of paths $[]$.
After the normalization, we get a prefix-free graph which can be directly represented in GFA format as shown in Figure \ref{fig:gfa}.

>>>>

\section{Prefix-free graphs}
The idea of prefix-free graphs is inspired by a technique used in the tool rsync named Context-Triggered Piecewise Hashing (CTPH) \cite{kornblum2006identifying}.
CTPH uses a rolling hash to partition a string into substrings such that long repeated substrings are partitioned the same way.
These substrings are then hashed with a traditional hash function and stored as a string signature.
The signatures of several files are then compared to determine changes.

In prefix-free graphs, we partition each sequence in a given pangenome into \emph{segment}s.
These segments form nodes of the prefix-free graph, and their adjacencies in the original sequences constitute edges.
The sequences are represented as \emph{path}s in the graph.

The segments have two essential characteristics making them a good choice for nodes of a pangenomic graph.
Similarly to CTPH, long repeated sequences will be partitioned the same way.
Furthermore, no segment is a prefix of another, making a set of segments prefix-free.
The second characteristic is crucial for connecting prefix-free graphs and suffix arrays, as will be presented in the next section.

@@

To illustrate the entire procedure, consider a set of sequences $P = \{$\texttt{CACGTACT}, \texttt{CACACT}, \texttt{CACGACT}$\}$ and a set of trigger words $T = \{$\texttt{AC}, \texttt{CG}$\}$.
After the partitioning, we obtain a set of segments with IDs $\{$\texttt{0:CAC}, \texttt{1:ACG}, \texttt{2:CGTAC}, \texttt{3:ACT..}, \texttt{4:ACAC}, \texttt{5:CGAC}$\}$ and a list of paths \texttt{[[0,1,2,3], [0,4,3], [0,1,5,3]]}.
After the normalization, we get a prefix-free graph which can be directly represented in GFA format as shown in Figure \ref{fig:gfa}.
